---
title: "Harris County Criminal Court and Race"
author: "Malsi Zambrano-Welch"
date: "2022-08-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(highcharter)
library(jastyle)
```

## WRU in the Harris County Criminal Court System

I explore the application of the WRU package (Imai & Khanna, 2016) on the Harris County Criminal Court data by offender. For each case, the offender is given an arresting-officer-determined race. The officer identifies the offender into one of five categories: white, black, asian, Native American, and unknown. The options do not included Hispanic/Latino. I explore the utility of the WRU package to add the probability an offender is Hispanic/Latino to the case information and explicitly use those predictions to evaluate differences in rates of offense category charges among offenders.

The wru package works by Bayesian probability prediction. Most specifically, it calculates the probability Race is equal to race (white, black, hispanic, asian, other) given the Surname and Geolocation.

Pr(Race|(Surname and Geolocation))

In order to calculate this, the wru package uses Census data including the racial composition of frequently occuring surnames (Census Surname List), and the racial composition of Census geographic divisions. 

##Criminal Court Data
The data itself is robust and full of avenues for exploration. I concentrate my exploration on rates of offense charge category by race and ethnicity.
```{r data, echo = FALSE}
data %>% pull(age) %>% 
  hchart(name = "Age of Offenders") %>% 
  hc_title(text = "Most offenders are in their mid-twenties and younger") %>% 
  hc_subtitle(text = "Distribution of Offender Age in Harris County Criminal Court") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_yAxis(title = "Number of Offenders", max = 20000) %>% 
  hc_xAxis(title = "Age of Offenders") %>% 
  hc_theme(ja_hc_theme())

full_filtered %>% select(def_sex) %>% tibble() %>% group_by(def_sex) %>% summarise(n = n()) %>% 
  filter(def_sex == "F" | def_sex == "M") %>% 
  hchart("bar", hcaes(x = def_sex, y = n)) %>% 
  hc_title(text = "The majority of offenders are male") %>% 
  hc_subtitle(text = "Distribution of Offender Age in Harris County Criminal Court") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_yAxis(title = "Number of Offenders") %>% 
  hc_xAxis(title = "Sex of Offenders") %>% 
  hc_theme(ja_hc_theme())

full_filtered %>% 
  select(ageatfiling, def_sex) %>% tibble() %>% 
  group_by(ageatfiling, def_sex) %>% 
  summarise(n = n()) %>% 
  filter(def_sex == "F" | def_sex == "M") %>% filter(is.na(ageatfiling) == FALSE) %>% 
hchart("column", hcaes(x = ageatfiling, y = n), stacking = "normal") %>% 
  hc_title(text = "The majority of offenders are young males at or below their midtwenties") %>% 
  hc_subtitle(text = "Distribution of Offender Age by Sex in Harris County Criminal Court") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_yAxis(title = list(text = "Number of Offenders", max = 20000)) %>% 
  hc_xAxis(title = list(text = "Age of Offenders")) %>% 
  hc_theme(ja_hc_theme())

full_filtered %>% select(def_rac, def_sex) %>% tibble() %>% 
  group_by(def_rac,def_sex) %>% summarise(n = n()) %>% 
  filter(def_sex == "F" | def_sex == "M") %>% 
  hchart("bar", hcaes(x = def_rac, y = n), stacking = "normal") %>% 
  hc_title(text = "The majority of offenders are males classified as white") %>% 
  hc_subtitle(text = "Distribution of Offender Race in Harris County Criminal Court") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_yAxis(title = "Number of Offenders") %>% 
  hc_xAxis(title = "Race of Offenders") %>% 
  hc_theme(ja_hc_theme())
```
The majority of offenders are white males and the age of offenders skews toward offenders in their early to mid-twenties. The WRU package is limited in its ability to utilize sex and age as predictive elements. I detail the characteristics of the dataset here to provide context

```{r, echo = FALSE}
### top charges using HCo Data ####

#can use either _wru as we are not yet getting into weights and apps of those numbers
#top ten off_cat
genpop_offcat <- name_wru %>% 
  group_by(offcat) %>% 
  summarise(n = n()) %>% arrange(desc(n))
genpop_offcat <- genpop_offcat %>% 
  mutate(percent = n/sum(n)*100) %>% 
  mutate(label = sub(".^","chr%",round(percent,0)))
genpop_offcat %>% slice(1:10) %>% 
  hchart("bar",hcaes(x = offcat, y = percent)) %>% 
  hc_title(text = "More offenders are charged with non-sexual assault than other offense categories") %>% 
  hc_subtitle(text = "Top Ten Offense Categories in Harris County Criminal Court, 2014-2019") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_yAxis(title = list(text = "% of cases receiving offense category charge")) %>%
  hc_xAxis(title = list(text = "Offense Category")) %>% 
  hc_plotOptions(
    bar = list(
      dataLabels = list(
        enabled = TRUE,
        formatter = JS(
          "
        function(){return(this.point.label)}
       "
        )
      )
    )
  ) %>% 
  hc_add_theme(ja_hc_theme()) 

#top ten by off_cat and race
genpop_offcat_race <- name_wru %>% 
  group_by(offcat,race) %>% 
  summarise(n = n()) %>% arrange(desc(n)) %>% ungroup()  %>% 
  mutate(race = case_when(race == 1 ~ "White",
                          race == 2 ~ "Black",
                          race == 4 ~ "Asian",
                          race == 5 ~ "Native American/Unknown"))
total_by_race <- genpop_offcat_race %>% 
  group_by(race) %>% 
  summarise(total = sum(n))
genpop_offcat_race <- genpop_offcat_race %>% 
  left_join(total_by_race, by = "race") %>% 
  mutate(percent = n/total * 100) %>% 
  mutate(label = sub(".^","chr%",round(percent,0))) %>% 
  group_by(race) %>% arrange(desc(n), .by_group = T)
  
genpop_offcat_race %>% filter(n > mean(n)) %>% 
  hchart("bar",hcaes(x = race,y = percent, group = offcat)) %>% 
  hc_title(text = "Prostitution and Unlawful Possession/Conduct with a Weapon are charge categories almost unique to asian and black offenders respectively") %>%
  hc_subtitle(text = "Share of Offense Categories by Police-Defined Race, categories greater than racial group average") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_xAxis(title = "Police-Defined Race") %>% 
  hc_yAxis(title = "% offense category", max = 20) %>% 
  hc_plotOptions(
    bar = list(
      dataLabels = list(
        enabled = TRUE,
        formatter = JS(
          "
        function(){return(this.point.label)}
       "
        )
      )
    )
  )   %>% 
  hc_add_theme(ja_hc_theme())
```
Black offenders are charged with unlawful possession/conduct with a weapon a higher rates than other races. They are also charged with nonsexual assault at greater rates than other racial groups, but the disparity is a small percentage point difference (+5% Asian, +2% Native American/unknown, +3% white).

## Application of wru Prediction to Each Case Offender

The wru prediction calculates a probability each case offender is of each race in the set: White, Black, Hispanic, Asian, Other. The random sample below demonstrates the calcuation. In this case, I show:
Pr(Race|Geolocation)
for three random cases.
```{r wru prediction}
name_wru %>% select(pred.whi,pred.bla,pred.his,pred.asi,pred.oth) %>% slice_sample(n=3)
```

I utilized these predictions in two different ways. The first is division of each case into a weighted entry by race. Here I show
Pr(Race|(Surname and Geolocation))

```{r wru pivot}
geoname_wru_weight %>% select(cas,pred_race,predict_value) %>% slice(7:11)
```
The second is using Hispanic/Latino as an ethnicity rather than race. This involves collapsing the predictions for white, black, asian, and other into a prediction for non-Hispanic/latino. Then, attaching ethnicity predictions to officer-defined racial categories. Each case has a defined race and a probability of ethnicity.
```{r wru ethnicity pivot}
geoname_wru_weight %>% select(cas,race,pred_race,predict_value) %>% filter(pred_race == "his" | pred_race == "nonhisp") %>% slice(1:4)

```

## Utlizing wru Predictions: Top Offense Categories by Race

First, I compare the top offense categories by predicted race given surname.

```{r top charges by race|sur using WRU results, echo=FALSE}
#### top charges by race using WRU results ####
#use pred_value as weights, group by pred_race and off_cat, sum on pred_value
weighted_offcat_pred_race_name <- name_wru_weight %>% filter(pred_race != "nonhisp") %>% 
  group_by(offcat, pred_race) %>% 
  summarise(n = sum(predict_value)) %>% arrange(desc(n)) %>% ungroup()

total_by_race <- weighted_offcat_pred_race_name %>% 
  group_by(pred_race) %>% 
  summarise(total = sum(n))
weighted_offcat_pred_race_name <- weighted_offcat_pred_race_name %>% 
  left_join(total_by_race, by = "pred_race") %>% 
  mutate(percent = n/total * 100) %>% 
  mutate(label = sub(".^","chr%",round(percent,0))) %>% 
  group_by(pred_race)

weighted_offcat_pred_race_name %>% filter(n > mean(n)) %>% 
  hchart("bar",hcaes(x = pred_race,y = percent, group = offcat)) %>% 
  hc_title(text = "Census name prediction suggests Hispanics charged under Alcohol-Driving comprise the largest offense category") %>%
  hc_subtitle(text = "Share of Offense Categories by Race Predicted by Name") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_xAxis(title = "Prediced Race by Name") %>% 
  hc_yAxis(title = "% offense category") %>% 
  hc_plotOptions(
    bar = list(
      dataLabels = list(
        enabled = TRUE,
        formatter = JS(
          "
        function(){return(this.point.label)}
       "
        )
      )
    )
  )   %>% 
  hc_add_theme(ja_hc_theme())

```
Then, I look at race given surname and geolocation and those offense category distributions.

```{r top charges by race|sur & geolocation using WRU results, echo=FALSE}
#### race|sur & geolocation ####
weighted_offcat_pred_race_geoname <- geoname_wru_weight %>% filter(pred_race != "nonhisp") %>% 
  group_by(offcat, pred_race) %>% 
  summarise(n = sum(predict_value)) %>% arrange(desc(n)) %>% ungroup()
total_by_race <- weighted_offcat_pred_race_geoname %>% 
  group_by(pred_race) %>% 
  summarise(total = sum(n))
weighted_offcat_pred_race_geoname <- weighted_offcat_pred_race_geoname %>% 
  left_join(total_by_race, by = "pred_race") %>% 
  mutate(percent = n/total * 100) %>% 
  mutate(label = sub(".^","chr%",round(percent,0))) %>% 
  group_by(pred_race)

weighted_offcat_pred_race_geoname %>% filter(n > mean(n)) %>% 
  hchart("bar",hcaes(x = pred_race,y = percent, group = offcat)) %>% 
  hc_title(text = "Geographic and Name-base prediction maintains Hispanics charged under Alcohol-Driving comprise the largest offense category") %>%
  hc_subtitle(text = "Share of Offense Categories by Predicted Race Given Name and Geolocation") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_xAxis(title = "Prediced Race") %>% 
  hc_yAxis(title = "% offense category") %>% 
  hc_plotOptions(
    bar = list(
      dataLabels = list(
        enabled = TRUE,
        formatter = JS(
          "
        function(){return(this.point.label)}
       "
        )
      )
    )
  )   %>% 
  hc_add_theme(ja_hc_theme())

```
Next, I evaluate the charge rate solely by ethnicity.

```{r ethnicity prediction and charge categories}
#### top charges by ethnicity ####
#use pred_value as weights, group by pred_ETHNICITY and off_cat, sum on pred_value
weighted_offcat_pred_eth_geoname <- geoname_wru_weight %>% 
  filter(pred_race == "nonhisp" | pred_race == "his") %>% 
  group_by(offcat, pred_race) %>% 
  summarise(n = sum(predict_value)) %>% arrange(desc(n)) %>% ungroup()
total_by_eth <- weighted_offcat_pred_eth_geoname %>% 
  group_by(pred_race) %>% 
  summarise(total = sum(n))
weighted_offcat_pred_eth_geoname <- weighted_offcat_pred_eth_geoname %>% 
  left_join(total_by_eth, by = "pred_race") %>% 
  mutate(percent = n/total * 100) %>% 
  mutate(label = sub(".^","chr%",round(percent,0))) %>% 
  group_by(pred_race)

weighted_offcat_pred_eth_geoname %>% filter(n > mean(n)) %>% 
  hchart("bar",hcaes(x = pred_race,y = percent, group = offcat)) %>% 
  hc_title(text = "Geographic and Name-base prediction maintains Hispanics charged under Alcohol-Driving comprise the largest offense category") %>%
  hc_subtitle(text = "Share of Offense Categories by Predicted Ethnicity Given Name and Geolocation") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_xAxis(title = "Prediced Ethnicity") %>% 
  hc_yAxis(title = "% offense category") %>% 
  hc_plotOptions(
    bar = list(
      dataLabels = list(
        enabled = TRUE,
        formatter = JS(
          "
        function(){return(this.point.label)}
       "
        )
      )
    )
  )   %>% 
  hc_add_theme(ja_hc_theme())
### interested to add predicted ethnicity to officer-defined race
  


weighted_offcat_pred_eth_name <- name_wru_weight %>% filter(pred_race == "nonhisp" | pred_race == "his") %>% 
  group_by(offcat, pred_race) %>% 
  summarise(n = sum(predict_value)) %>% arrange(desc(n))%>% ungroup()
total_by_eth <- weighted_offcat_pred_eth_name %>% 
  group_by(pred_race) %>% 
  summarise(total = sum(n))
weighted_offcat_pred_eth_name <- weighted_offcat_pred_eth_name %>% 
  left_join(total_by_eth, by = "pred_race") %>% 
  mutate(percent = n/total * 100) %>% 
  mutate(label = sub(".^","chr%",round(percent,0))) %>% 
  group_by(pred_race)

weighted_offcat_pred_eth_name %>% filter(n > mean(n)) %>% 
  hchart("bar",hcaes(x = pred_race,y = percent, group = offcat)) %>% 
  hc_title(text = "Name-based prediction maintains Hispanics charged under Alcohol-Driving comprise the largest offense category") %>%
  hc_subtitle(text = "Share of Offense Categories by Predicted Ethnicity Given Name") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_xAxis(title = "Prediced Ethnicity") %>% 
  hc_yAxis(title = "% offense category") %>% 
  hc_plotOptions(
    bar = list(
      dataLabels = list(
        enabled = TRUE,
        formatter = JS(
          "
        function(){return(this.point.label)}
       "
        )
      )
    )
  )   %>% 
  hc_add_theme(ja_hc_theme())
```
Finally, we evaluate by adding ethnicity to the office-defined race.
```{r}

#### top charges by race and predicted ethnicity ####
weighted_offcat_pred_race_eth_name <- name_wru_weight %>% 
  filter(pred_race == "nonhisp" | pred_race == "his") %>% 
  group_by(offcat, race, pred_race) %>% 
  summarise(n = sum(predict_value)) %>% arrange(desc(n)) %>% ungroup()

weighted_offcat_pred_race_eth_name %>% filter(n > 6000) %>% 
  hchart("bar", hcaes(x = offcat, y = n, group = pred_race)) %>% 
  hc_title(text = "Non-Hispanic Blacks charged with non-sexual assault more often than any other racial-ethnic group") %>%
  hc_subtitle(text = "Number of Offense Category Charges by Race and Predicted Ethnicity") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_xAxis(title = list(text = "Offense Category")) %>% 
  hc_yAxis(title = list(text = "# of cases")) %>% 
  hc_add_theme(ja_hc_theme())
##non-hisp whites are not in the top five charge categories, begin at 6 with Alcohol-Driving
### by numbers, but the grouping/stacking isn't quite right...

total_race_eth <- weighted_offcat_pred_race_eth_name %>% 
  group_by(race, pred_race) %>% 
  summarise(total = sum(n))
weighted_offcat_pred_race_eth_name <- weighted_offcat_pred_race_eth_name %>% 
  left_join(total_race_eth, by = c("race", "pred_race")) %>% 
  mutate(percent = n/total * 100) %>% 
  mutate(label = sub(".^","chr%",round(percent,0))) %>% 
  group_by(pred_race)
weighted_offcat_pred_race_eth_name %>% filter(n > 6000) %>% 
  hchart("bar", hcaes(x = offcat, y = n),stacking = "normal") %>% 
  hc_title(text = "Non-Hispanic Blacks charged with non-sexual assault more often than any other racial-ethnic group") %>%
  hc_subtitle(text = "Rate of Offense Category Charge by Race and Predicted Ethnicity") %>% 
  hc_caption(text = "Harris County Criminal Court, 2014-2019") %>% 
  hc_xAxis(title = list(text = "Offense Category")) %>% 
  hc_yAxis(title = list(text = "% of cases")) %>% 
  hc_add_theme(ja_hc_theme())  %>% 
  hc_plotOptions(
    bar = list(
      dataLabels = list(
        enabled = TRUE,
        formatter = JS(
          "
        function(){return(this.point.label)}
       "
        )
      )
    )
  )   %>% 
  hc_add_theme(ja_hc_theme())

```

